---
title: "Health vs. Income"
author: "Yiyun Gong"
output: html_notebook
---

```{r message=FALSE, warning=FALSE}
# always clean up R environment
rm(list = ls())

# load all packages here
library(dplyr)
library(reshape2)
library(tidyr)
library(ggplot2)
library(mosaic)
library(mdsr) 
library(Lahman)
library(readxl)
library(ape)
library(mclust)
library(lubridate)
library(NHANES)
library(glmnet)
library(rpart)
library(rpart.plot)
library(randomForest)
library(nnet)
library(e1071)
library(class)
library(rsample)
library(ipred)
library(caret)
library(partykit)
library(tibble)
library(gbm)
# load dataset
health_allPop<-
  read.csv("Health_allPop.csv", header = T)
health_percapita<-
  read.csv("Health_percapita.csv", header = T)
income<-
  read.csv("income.csv", header = T)
income_num<-
  read.csv("income_num.csv", header = T)
# user-defined function
remove_comma<-
  function(table_name){
    ncol<-
      ncol(table_name)
    for (i in 2:ncol){
      table_name[,i]<-
        gsub(",","",table_name[,i])
      table_name[,i]<-
        as.numeric(table_name[,i])
    }
    return(table_name)
  }

#user-defined simulation function
hein<-
  function(a, b, c, d, e){
    health_service<-
      sample(a, 1)
    health_equipment<-
      sample(b, 1)
    medianIncome<-
      sample(c, 1)
    lowIncome<-
      sample(d, 1)
    highIncome<-
      sample(e, 1)
    health_sim<-
      health_service+health_equipment
    percent_med<-
      health_sim/medianIncome
    percent_low<-
      health_sim/lowIncome
    percent_high<-
      health_sim/highIncome
      
    return(data.frame(health_sim, health_service, health_equipment, 
                      medianIncome, percent_med, lowIncome, percent_low, 
                      highIncome,  percent_high))
  }
```

#Data Wrangling

####Data Cleaning income_num dataset
```{r}
#Clean column year
income_num$Year<-
  substr(income_num$Year, 1, 4)
#Delete repeated year data
income_num<-
  income_num[-5,]
#Change the Year column to numeric number
income_num$Year<-
  as.numeric(income_num$Year)
```

```{r}
#remove_comman function at the user-defined function section
#Remove all the comman from the whole dataset
income_num<-
  remove_comma(income_num)
```

####Subset and merge datasets income_num and health_percapita
```{r}
income_num_sub<-
  income_num %>%
  filter(Year>1999&Year<2016)%>%
  arrange(Year)
```

```{r}
#remove commas from dataset health_percapita and health_allPop
health_percapita<-
  remove_comma(health_percapita)
health_allPop<-
  remove_comma(health_allPop)
```

####Join Income table with Average insurance price per capita table
Two meaningful tables which are health_percapita table, health care expenditures on each sub section per capita and income_num_sub table, Mean Income for each fifth percent from 2000-2015 are joined together to be further analyze the how important health care expenditures take places on Income.
```{r}
insurance_income<-
  merge(x=income_num_sub, y=health_percapita, by = "Year", all=TRUE)
```

####Calculate the percentage of insurance/median income
```{r}
insurance_income<-
  insurance_income %>%
  mutate(ins_percentage = Health/Third.fifth)
```

####Spread Income Percentage Table by Subject and Family Status
```{r}
income_status<-
  income %>%
  spread(key = Subject, value = Estimate)
```

```{r}
income_subject<-
  income %>%
  spread(key = Status, value = Estimate)
```

####Find the largest percentage on Subject by each Family Status per Year
```{r}
income_status_sub<-
  income_status[,3:12]
income_status$max<-
  colnames(income_status_sub)[apply(income_status_sub, 1, which.max)]
income_status$maxPerc<-
  apply(income_status_sub, 1, max)
head(income_status[,c(1,2,13,14)], 5)
```

####Apply function to calculate the median on Health_percapita Table
```{r}
health_percapita %>%
  select(Health, Health.services, Medical.products..appliances.and.equipment) %>%
  apply(MARGIN = 2, FUN = median, na.rm=TRUE)
```

#Data Visualization

####US Income Distribution
**Graph that display four variables in the same graph**
```{r fig.width=5, fig.asp=0.4}
#US Income distribution
ggplot(data = income_num, aes(x = Year, y=Third.fifth, colour="Median Income"))+
  geom_line()+
  geom_line(data = income_num, aes(x=Year, y=Lowest.fifth, colour="Lowest fifth mean income"))+
  geom_line(data = income_num, aes(x=Year, y=Highest.fifth, colour="Highest fifth mean income"))+
  geom_line(data = income_num, aes(x=Year, y=Top.5.percent, colour="Top 5% mean income"))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))+
  xlab("Year")+
  ylab("Income")+
  scale_x_continuous(breaks = income_num$Year)+
  ggtitle("US Income Distribution over Years")
```

####Health Expenditures and Median Income 2000-2015
**Graph with layered data from secondary data frame**
```{r fig.width=5, fig.asp=0.7, message=FALSE}
#need to change the labels and titles
ggplot(data = health_percapita, aes(x = Year, y = Health)) + 
  geom_bar(stat = "identity") +
  geom_point(data = income_num_sub, aes(x=Year, y=Third.fifth))+
  geom_smooth(data = income_num_sub, aes(x=Year, y=Third.fifth), show.legend = FALSE, method = "auto")+
  ggtitle("Health Expenditures and Median Income 2000-2015") +
  ylab("Health expenses and Income")+
  scale_x_continuous(breaks = insurance_income$Year)
```

####Percent of Insurance Expenses over Income per year
```{r message=FALSE, warning=FALSE}
ggplot(data = insurance_income, aes(x = Year, y = ins_percentage))+
  geom_point()+
  geom_smooth(method = "lm")+
  ggtitle("Percent of Insurance Expenses over Income per year")+
  scale_x_continuous(breaks = insurance_income$Year)
```

####Median Income spread versus Health Services Expense over the Years
```{r message=FALSE, warning=FALSE}
ggplot(data = insurance_income,
            aes(x = Third.fifth, y = Health.services)) +
  geom_bar(fill = "grey", stat = "identity") +
  ylab("Average Insurance Charges ($)") + xlab("Average Income") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Median Income versus Health Services Expense")
```

####Income Spread based on Subject and Family Status
```{r fig.width=5, fig.asp=0.7, message=FALSE}
ggplot(data = income, aes(x = Estimate, fill=Subject))+
  geom_density()+
  xlab("Percentage in Each Subject")+
  ggtitle("Income Spread based on Subject and Family Status")
```

#Data Analysis

#Statistical modeling/Supervised learning
## Data Preparation
```{r}
# set RNG seed for reproducible results
set.seed(688) 
# partition the data
n <- 
  nrow(income_status)
test_idx <- 
  sample.int(n, size = round(0.2 * n)) 
# select row numbers for the test set
train <- 
  income_status[-test_idx, ]  
# exclude the test set cases
nrow(train)
```

```{r}
test <- 
  income_status[test_idx, ]    # test set cases only
nrow(test)
```

## Inspecting training data
```{r}
tally(~ max, data = train)
```

```{r}
tally(~ Status, data=train)
favstats(train$`$100,000 to $149,999`)
favstats(train$`$15,000 to $24,999`)
favstats(train$`$50,000 to $74,999`)
favstats(train$maxPerc)
```

```{r fig.width=5, fig.asp=0.7}
pairs(train[, c(2:12)])
```

## Null Model
```{r}
mod_null <- 
  tally(~ max, data = train, format = "percent")
mod_null
```


#Logistic Regression
```{r warning=FALSE, message=FALSE}
predictors<- 
  model.matrix(maxPerc~., data = train) 
fit<- 
  cv.glmnet(predictors, train$maxPerc)
fit
```

```{r fig.width=6, fig.asp=0.5}
plot(fit)
```

```{r}
mean(fit$cvlo)
mean(fit$cvup)
```

## Decision Trees
```{r}
rpart(max ~ `$50,000 to $74,999`, data = train)
```

```{r}
split <- 17.85 # first split from simple `rpart`
train %>%
  ggplot(aes(x = `$50,000 to $74,999`, y = max)) + 
  geom_point(aes(color = 'Year'), 
             position = position_jitter(width = 0, height = 0.15), 
             alpha = 0.4) + 
  geom_vline(xintercept = split, color = "blue", lty = 2)+
  scale_x_continuous(breaks = c(1:20))
```

```{r}
mod_tree <- 
  rpart(max~., data = train, control = rpart.control(cp=0.03))
mod_tree
```

```{r fig.width=5, fig.asp=0.7}
plot(as.party(mod_tree))
```

```{r}
train_tree2 <- 
  train %>%
  mutate(sub_dtree = predict(mod_tree, type = "class"))
confusion <- 
  tally(sub_dtree ~ max, data = train_tree2, format = "count")
confusion
remove(train_tree2)
```

```{r}
sum(diag(confusion)) / nrow(train) * 100
```

#Unsupervised learning

##Hierarchical clustering
```{r}
unsuperSet<-
  insurance_income
rownames(unsuperSet)<-
  c(2000:2015)
```

```{r}
set_diffs<-
  dist(unsuperSet)
str(set_diffs)
```

```{r}
set_mat<-
  set_diffs %>% 
  as.matrix() %>%
  round(digits = 2)
#glimpse on the matrix
set_mat[1:6, 1:6]
```

```{r}
set_diffs %>%
  hclust() %>%
  as.phylo() %>%
  plot(cex=0.9, label.offset = 1)
```
A dendrogram constructed by hierarchical clustering from year-to-year distance data.

##K-means
```{r}
supSet<-
  insurance_income
set.seed(2500)
set_clusts_ii<- 
  supSet %>% 
  kmeans(centers = 8)
str(set_clusts_ii)
```


```{r fig.width=5, fig.asp=0.7}
set.seed(2500)
set_clusts_ii<- 
  supSet %>% 
  kmeans(centers = 8) %>%
  fitted("classes") %>% 
  as.character()
supSet <- 
  supSet %>% mutate(cluster = set_clusts_ii)
supSet %>% 
  ggplot(aes(x = Year, y = ins_percentage)) +
  geom_point(aes(color = cluster), alpha = 0.5)+
  scale_x_continuous(breaks = income_num$Year)
```

```{r}
supSet<-
  income[,c('Year', 'Estimate')]
set.seed(1200)
set_clusts <- 
  supSet %>% 
  kmeans(centers = 8)
str(set_clusts)
```

```{r}
set.seed(1200)
set_clusts <- 
  supSet %>% 
  kmeans(centers = 8) %>%
  fitted("classes") %>% 
  as.character()
supSet <- 
  supSet %>% mutate(cluster = set_clusts)
supSet %>% 
  ggplot(aes(x = Year, y = Estimate)) +
  geom_point(aes(color = cluster), alpha = 0.5)+
  scale_x_continuous(breaks = income$Year)
```

#User-defined simulation(Bootstrapping)
```{r}
set.seed(800)
healthserv_set<-
  insurance_income$Health.services
equip_set<-
  insurance_income$Medical.products..appliances.and.equipment
income_set<-
  insurance_income$Third.fifth
income_top<-
  insurance_income$Highest.fifth
income_low<-
  insurance_income$Lowest.fifth
percent_sim<-
  do(5000)*hein(healthserv_set, equip_set, income_set, income_low, income_top)
```

```{r}
head(percent_sim)
```

```{r}
summary(percent_sim$health_sim)
summary(percent_sim$health_service)
summary(percent_sim$health_equipment)
```

```{r fig.width=5, fig.asp=0.7}
ggplot(data=percent_sim, aes(x=health_sim, color="Health Expenditures"))+
  geom_density()+
  geom_density(data=percent_sim, aes(x=medianIncome, color="medianIncome"))+
  geom_density(data=percent_sim, aes(x=highIncome, color="highIncome"))+
  geom_density(data=percent_sim, aes(x=lowIncome, color="lowIncome"))+
  scale_x_continuous("Expenditures on insurance and income levels")
```
```{r}
#ACA affordability percentage
#data from website:
afford<-
  mean(c(0.0956, 0.0966, 0.0969, 0.0956, 0.0986))
afford
```


```{r fig.width=3, fig.asp=0.6}
ggplot(data=percent_sim, aes(x=percent_med, color="median income"))+
  geom_density()+
  geom_density(data=percent_sim, aes(x=percent_low, color="Lowest fifth"))+
  geom_density(data=percent_sim, aes(x=percent_high, color="Highest fifth"))+
  geom_vline(xintercept = afford, linetype = 2, color="Brown")+
  scale_x_continuous("percent of health expenditures on income", breaks = (0:10)/10)
```





